---
title: "Homework 1"
author: "Enter your name here"
output: html_document
---

## R Basics

You can insert an R code chunk into an R Markdown document by using a keyboard shortcut. The default keyboard shortcut to insert an R code is:

Windows/Linux: Ctrl + Alt + I
Mac: Command + Option + I

Make sure to comment on your code using #

#### Problem 1:

1. From set = {1.5, 15, 0, 54, 23}, pick two numbers and perform the Following Operations:

.Subtraction
.Multiplication
.square root

```{r}

```

#### Problem 2

Suppose you have a set of numbers representing the assignments scores:

```{r}
scores <- c(87,100, 91, 95, 81.5, 0, 39, 74, 92)

#Calculate the average of the scores, and round your answer to two decimal number


#Find the Maximum and Minimum score


#Find how many scores were above the average, use coding for this


#Create a new set called new_scores that includes the elements from indices 2 to 6 of the scores vector

```



#### Problem 3: 

Read section 2.3, then answer the following questions

```{r}
# Create a vector with the number 1 repeated 5 times. Print it


# Create a vector with a sequence of numbers from 1 to 10 with a step of 2. Print it


# Create a vector by repeating a sequence from 1 to 3, 2 times. Print it.


```



#### Problem 4

```{r}

mixed_vector <- c(1, "two", 3.0, TRUE, 2)

# Check the class of the vector

## Coerce to numeric and print it 


```



