---
title: "Homework 5"
author: "Raul Miranda"
date: 20 Oct 2025
output: html_document
---

#### Setup
Install and load the necessary packages:

```{r}

library(tidyverse)
library(dplyr)
data("mtcars")
```

#### Functions

1- Create a function named calculate_area_of_circle that takes one parameter, radius, representing the radius of a circle. The function should calculate and return the area using the formula:

area = ğœ‹ Ã— radius^2

Test your function with at least two different radii and print the results.

```{r}

calculate_area_of_circle <- function (r) {
     if (r<0) {return ("Give me a non-negative value please....")}
     area <- pi * r^2
   return (area)
}

print (calculate_area_of_circle (5))

print (calculate_area_of_circle (10))

print (calculate_area_of_circle (0))

print (calculate_area_of_circle (-10))
```



2- Write a function called check_temperature that takes a single numeric input temp.

If temp > 30, print "Hot"

If temp >= 15 & temp <= 30, print "Warm"

Otherwise, print "Cold"

Test the function on three different values (e.g., 10, 20, 35).

```{r}

check_temperature <- function (t) {
  case_when(t>30 ~ 'Hot', t>=15 & t<=30 ~ 'Warm', .default='Cold')
}

check_temperature (10)
check_temperature (20)
check_temperature (35)
check_temperature (-5)


# for practice, lets use if else

check_temperature2 <- function (t) {
  if (t>30) {return ('Hot')}
  if (t>=15 & t<=30) {return ('Warm')}
  else {return ('Cold')}
} 

check_temperature2 (10)
check_temperature2 (20)
check_temperature2 (35)
check_temperature2 (-5)
```


3- Write a function named sum_odd_numbers that uses a for loop to calculate the sum of all even numbers from 1 to n. The function should return the total sum.  

#### note: I interpret that the function should be called  sum_even_numbers

Test the function with n = 9 and n = 39.


```{r}
# for even numbers:

sum_even_numbers <- function (n) {
    tot <- 0
    upto <- n%/%2
    for (i in 1:upto) {
      tot <- tot + 2*i
    }
    return (tot)
  }

sum_even_numbers(9)
sum_even_numbers(39)


```

```{r}
# the following is for practice -


# using if 

sum_even_numbers <- function (n) {
    tot <- 0
    for (i in 1:n) {
      if (i %% 2 == 0) { tot <- tot + i}
    }
    return (tot)
}

sum_even_numbers(9)
sum_even_numbers(10)

sum_odd_numbers <- function (n) {
    tot <- 0
    for (i in 1:n) {
      if (i %% 2 == 1) { tot <- tot + i}
    }
    return (tot)
}

sum_odd_numbers(9)
sum_odd_numbers(10)


# play using remainders

sum_odd_numbers <- function (n) {
    tot <- 0
    remainder <- case_when (n%%2==0 ~0, n%%2!=0 ~1)
    upto <- n%/%2 + remainder
    for (i in 1:upto) {
      tot <- tot + 2*i -1
    }
    return (tot)
  }

sum_odd_numbers(9)
sum_odd_numbers(10)
sum_odd_numbers(11)
sum_odd_numbers(12)

```




#### EDA

For this section, use the built-in mtcars dataset (instead of airquality). Perform the following:

- Check the structure of the dataset.

- Display the summary statistics.

- Check for any missing values.

Ask a simple question that can be answered with a basic function (e.g., What is the mean miles per gallon (mpg) across all cars?). Write the code and output the answer.

```{r}
mtcars               # show the dataframe
str (mtcars)         # show structure
summary(mtcars)      # get basic stats
colSums(is.na(mtcars))    # show count of missing values in each column -- 0


# question: what are the mean and median hp across all cars?
  
print ( mean (mtcars$hp) )                    # dataset has no NAs, so na.rm=TRUE unnecessary
print ( median (mtcars$hp) )                  # dataset has no NAs, so na.rm=TRUE unnecessary


# or using summarize

mtcars |>
summarize ( hp_mean = mean (hp),
            hp_median = median (hp))


```

### Published 

Published at: https://rpubs.com/rmiranda/1357863
